---
title: "i-Vectors in R"
author: "Avril Coghlan, adapted by Nathan Brouwer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This is a modification of ["DNA Sequence Statistics (2)"](https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter2.html) from Avril Coghlan's [*A little book of R for bioinformatics.*](https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/index.html).  Most of text and code was originally written by Dr. Coghlan and distributed under the [Creative Commons 3.0](https://creativecommons.org/licenses/by/3.0/us/) license. 



## Vectors in R

In a previous vignette, you learned about **variables** in R, such as **scalars**, **vectors**, and **lists**. You also learned how to use **functions** to carry out operations on variables, for example, using the log10() function to calculate the log to the base 10 of a scalar variable x, or using the mean() function to calculate the average of the values in a vector variable myvector.  For example, we can use log10() on a scalar like this:

```{r}
x <- 100
log10(x)
```

We can use mean() on a vector like this
```{r}
myvector <- c(30,16,303,99,11,111)
mean(myvector)
```
 
 
You also learned that you can extract an **element** of a vector by typing the vector name with the index of that element given in **square brackets**. For example, to get the value of the 3rd element in the vector myvector, we type:


```{r}
myvector[3]
```

A useful function in R is the seq() function, which can be used to create a sequence of numbers that run from a particular number to another particular number. For example, if we want to create the sequence of numbers from 1-10 in steps of 1 (ie. 1, 2, 3, 4, ... 10), we can type:

```{r}
seq(1, 10, by = 1)
```



We can change the **step size** by altering the value of the “by” argument given to the function seq(). For example, if we want to create a sequence of numbers from 1-100 in steps of 20 (ie. 1, 21, 41, ... 101), we can type:

```{r}
seq(1, 101, by = 20)
```



## Vectors can hold numeric or character data

The vector we created above holds numeric data

```{r}
is(myvector)
```

As we've seen, vectors can holder character data, like the genetic code

```{r}
myvector <- c("A","T","G")
```

## Regular expressions can modify character data

We can use regular expressions to modify character data. For example, change the Ts to Us

```{r}
myvector <- gsub("T","U",myvector)
```

Now check it out
```{r}
myvector
```



## Plotting vectors

R allows the production of a variety of plots, including scatterplots, histograms, piecharts, and boxplots. For example, if you have two vectors of numbers myvector1 and myvector2, you can plot a scatterplot of the values in myvector1 against the values in myvector2 using the plot() function. If you want to label the axes on the plot, you can do this by giving the plot() function values for its optional arguments xlab and ylab:

```{r}
myvector1 <- c(10, 15, 22, 35, 43)
myvector2 <- c(3, 3.2, 3.9, 4.1, 5.2)

plot(myvector1, myvector2, xlab="vector1", ylab="vector2")
```


Base R has lots of plotting functions; additionally, peole have written packages to implement new plotting capabilities. ggplot2 is currently the most popular plotting package, and ggpubr is a package which makes ggplot2 easier to use.  For quick plots we'll use base R functions, and when we get to more important things we'll use ggplot2 and ggpubr.

