---
title: "x) Creating this package"
author: "Nathan Brouwer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

The following code outlines the key steps I use to set up the framework for a package, principally using the usethis package.


## Packages

I use these packages for setting things up
```{r, eval = F}
library(devtools)
library(usethis)
library(here)
```

## Create repository

I create git repo via the GitHub website, the cloned into it by hand by creating an RStudio Project.

## Set up package with usethis

I then use usethis::create_package() within my project directory to build the basic package infrastructure  This overights the initial project (after a handy prompt in the console).
```{r eval = F}
usethis::create_package(path = getwd())
```

### Create vignette infrastructure

I create vignette infrastructure and dummy vignette
```{r, eval = F}
usethis::use_vignette("temp")
```

I then write my vignettes using this template and change the file name.  I use "a)", "b)", "c)" as a prefix on my vignettes so that show up in order when the website is rendered using packagedown.


### Create "readme" and "news" files

```{r, eval = F}
use_readme_md()
use_news_md()
```

### Populating the readme file

I add basic information to the readme fill and add sample code for a call to devtools::install_github() to install the package.

```{r, eval = F}
# install.packages("devtools")
devtools::install_github("brouwern/dayoff")
```



### Create git ignore

Am not sure how to use this properly yet : (

```{r, eval = F}
use_git_ignore(".pdf", directory = ".")
```


Add typical files to gitignore: '.Rproj.user', '.Rhistory', '.Rdata', '.DS_Store'
```{r}
usethis::git_vaccinate()
```


### Add required packages

these are the packages I typically use.  I should vectorize this so it would look nicer : )

```{r, eval = F}
usethis::use_package("dplyr",   "Imports")
usethis::use_package("tidyr",   "Imports")
usethis::use_package("ggplot2", "Imports")
usethis::use_package("ggpubr",  "Imports")

# use_package("nlme", "Imports")
# use_package("lme4", "Imports")
# use_package("bbmle", "Imports")
# use_package("RCurl", "Imports")
# use_package("GGally", "Imports")
# use_package("broom", "Imports")

# use_package("arm", "Imports")

# devtools::use_package("data.table", "Imports")
# devtools::use_package("car", "Imports")
# devtools::use_package("lawstat", "Imports")
# devtools::use_package("effsize", "Imports")
# devtools::use_package("here", "Imports")
# devtools::use_package("metafor", "Imports")
# devtools::use_package("Formula", "Imports")
# devtools::use_package("data.tree", "Imports")




```



### Don't save/load user workspace between sessions

```{r, eval = F}
usethis::use_blank_slate()
```


### Use roxygen for documentation

Where would we be without roxygen

```{r, eval = F}
usethis::use_roxygen_md()
```


### Package-level documents

"Adds a dummy .R file that will prompt roxygen to generate basic package-level documentation. "

```{r, eval = F}
usethis::use_package_doc()
```


### Use pkgdown

For making front end website

```{r, eval = F}
usethis::use_pkgdown()
```



## Set up data

Create folder for external data
```{r}
dir.create(here::here("/inst"))
dir.create(here::here("/inst/extdata"))
```

This could be done with use_directory()

### Look at data in my extdata file


External data is stored in "/inst/extdata"
```{r, eval = F}
list.files(here::here("/inst/extdata"))
```

#### Raw data

```{r eval = F}
# milk_raw <- read.csv(here::here("/inst/extdata",
#                     "skibiel_mammalsmilk_raw.csv"))
# 
# use_data(milk_raw)
# use_r("milk_raw")
```

```{r}
use_data_raw(name = "dengueseq", open = interactive())
write(dengueseq, file="dengueseq.fasta")
```


## License

For information on licenses see
http://kbroman.org/pkg_primer/pages/licenses.html

Plaintext versions of licenses can be found at
https://creativecommons.org/2011/04/15/plaintext-versions-of-creative-commons-licenses-and-cc0/

```{r, eval = F}
usethis::use_ccby_license(name = "Nathan Brouwer")
```






```{r}
#use_data_raw()
```

### Spell check

use_spell_check


### Other potentially useful usethis functions

use_r() #Create or edit a .R file


use_build_ignore()
Add files to .Rbuildignore


use_version() use_dev_version()
Increment package version



edit_r_profile() 
edit_r_environ() 
edit_r_makevars() 
edit_rstudio_snippets() 
edit_git_config() 
edit_git_ignore()




Use a directory




