---
title: "The Dayhoff PAM matrices"
author: "Nathan Brouwer"
date: "9/24/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

Alignments are ubiquitous in molecular biology, phylogenetics, and genomics.  For any two sequences, there are many ways you could align them by adding insertions or deletions.  The essential steps in alignment are therfore

* locate sections in the two sequences that are identical
* hypothesize the occurence of insertions and deletions that allow you to line up more of the sequence
* decide when there the amino acids or bases in the position are different due to mutation

Because there are so many ways to align two sequences there is a need to be able to judge how "good" a given alignment is a rank the quality of many alignments.  Key to this are alignment scoring matrices such as **BLOSUM** and **PAM** matrices.  The matrices are based on empirical data about how frequently differnt mutations have occurred between sequences.  

There are a number of steps required to calcualte these matrices.  The following tutorial will step you through key aspects of constructing Margaret Dayhoff's PAM matrices.  The focus here will be on the R operations need to reconstruct the matrices.  However, seeing how Dayhoff moved from raw data to a probablity matrix, and finally her scoring matrix should help you bettr understand how they work.


## Key steps

* Examine and finalize the set up of Dayhoff's data matrix (we'll call is pams)
* Set up Dayhoff's data on the relative frequeny of amino acids and the "**mutability**" of each of the 20 amino acids
* combine this information to make a **mutation probability matrix**


## Biology and data analysis Learning objectives

By the end of this exericse and after reviewing relevant readings you should be able to 

* define mutability
* define what a mutation probability matrix
* define what an alignment scoring matrix is
* better interpret histograms and scatterpots
* better interpret correlations

## Coding learning objectives

* practice building data frames by hand
* practice doing math with matrices and vectors
* introduction to for() loops


## Preliminaries

```{r}
library(dayoff)
library(ggpubr)
```


## Dayhoff's PAM matrix


The data in Figure 80 of Dayhoff et al (1978) is stored in the dataframe pams in the dayoff package.  Load this with the data() command. 
```{r}
# Make sure you've done libary(dayoff) !

data(pams)

```

How large is the data set?
```{r}



```


What kind of object is pams?
```{r}




```


Since its not a matrix we can't access the diagonal directly.  We can subset the dataframe and turn it into a matrix.  Take a look at the top of the dataframe using head()

```{r}


```

Check out what the different columns are using summary()
```{r}


```


The first column is different than all of the others.  You can check this using the is() function on just the column., eg is(pams$code) 


```{r}
   



```




We can remove this first column using R's square bracket notation.  Let's do this and save it to a new object called A. 

```{r}

A <- pams[ , -1]


```


Confirm the size of the dataframe using an appropriate command
```{r}


```


For some annoying reasons data.frames and matrices don't behave the same in R.  So if we want to work with these data as a matrix we need to convert it explicilty to a matrix.

```{r}

A <- as.matrix(A)

```

We can use the dayoff function tri_print() to look at the matrix
```{r}

tri_print(A)

```



Use the diag() function.  Why do you get this response?
```{r}



```


The matrix is currently in a **lower triangular** form.  The matrix will acctually be easier to work with when doing math if its in a **symmetric triangular** form.  You don't need to understand the following code, just run it. (For the curious: what it's doing is fidning all the "NA" values with is.na(), re-assigning, them to 0, then adding the original matrix and its mirror image (transpose)  to get a symetric matrix.

```{r}
# just run this code - no need to understand it
A[is.na(A)] <-0

A <- A + t(A)
```

Check out the new matrix.  Can you tell why I am calling it "symmetric" ?
```{r}
tri_print(A, triangle = F)
```


If its not clear why this is symmetric, run this coe, which will just show the first 4 rows and columns
```{r}

tri_print(A[1:4,1:4], triangle = F)

```



## Frequencies of amino acids in Dayhoff data

Dayhoff et al tallied up all of the amino acids in the data set they used and calculated the relative frequency of each one.  They shows the data in Table 22 (Table 3.1 in Pevsner).

We can recreate this datatable in. First, the frequencies.  
```{r}
frequencies <- c(0.089, 0.087, 0.085, 0.081, 0.070, 
                 0.065, 0.058, 0.051, 0.050, 0.047, 
                 0.041, 0.040, 0.040, 0.038, 0.037, 
                 0.034, 0.033, 0.030, 0.015, 0.010)

```

You should be able to determine what type of object this is, how long it is, etc.


Now the amino acids.

```{r}
aa <- c("Gly", "Ala", "Leu", "Lys", "Ser", 
        "Val", "Thr", "Pro", "Glu", "Asp",
        "Arg", "Asn", "Phe", "Gln", "Ile",
        "His", "Cys", "Tyr", "Met", "Trp")
```


Now use data.frame to turn this into a dataframe

```{r}

freqs <- data.frame(aa, frequencies) 


```


The code below will re-create the table as it is published.  This table is a little goofy from a data analysis perspective because there are two columns of amino acid names and two columns of frequencies.  First, I'm going to grab each side of the table.

```{r}
# first aa column and first frequency column
left  <- freqs[1:10, ]

# 2nd aa column and 2nd frequency column
right <- freqs[11:20, ]

```

Now I can put them together with a function called cbind(), which stands for "column bind"

```{r}

table22 <- cbind(left, right)

```




PUll up the table and compare it to the original.  Let me know if I made a mistake : )
```{r}

```



The table isn't handy for doing much but looking at.  For other stuff we need the original data in the freq object.  We can make a histogram of these frequencies using gghistogram().  Make sure ggpubr is loaded before you run this code (Note: you'll get an warnings; ignore this)


```{r}
#ignore the warning
gghistogram(data = freqs,  #use freqs, not table2
            x = "frequencies",
            bins = 15)
```

This is a histogram of **freuqncies**.  What does it mean that the tallest bar, is located between 0.04 and 0.04, goes up to 5?  Use the data table which amino acids are represented by this bar?  A function that can help us figure this out is which().  We can ask R, "which values are less than 0.04" like this.

```{r}
rows.less.04 <- which(freqs$frequencies < 0.04)
```

This gives us the row number of the values less than 0.04. 

```{r}
rows.less.04
```


We can look at the actual data like this
```{r}
freqs[rows.less.04, ]
```


## Table 21

The pams matrix above tells us about observed mutations.  Key to Dayhoff et al's calcualtions their PAM scoring matrices is the calculation of the realtive mutability of each amino acid.  Dayhoff et al explain it this way

> "A complete picture of the mutational process must include a consideration of the amino acids that did not change, as well as those that did.  For this we nee to know the probability that each amino acid will change in a given small evolutionary time interval.  We call this number the 'relative mutability'" (pag 346-347).

(See the Dayhoff chapter and pages 80-82 in Pevsner for more information).



Dayhoff doesn't provide all the data, just the final mutabilities.  The calculation is basically  (# of times the aa change)/(# of time the aa occurred).  

We can determine the number of times an amino acid occured like this

```{r}
sum(A[,"Ala"])
```

Or, equivalently
```{r}
sum(A[,1])
```
Or, because we made the matrix symmetric

```{r}
sum(A[1,])
```

If it seems weird that the last code wored, examine the results of this code, and look at the original matrix
```{r}
A[1,] == A[,1]
```


So, they observed Ala mutating to another amino acid 14576 times.  They then determined how many tiems Ala occured overall, did a little math to make everything work they way they wanted, and produced table 21.  Note that they've standardized things so that Ala is set as a reference point.


We can build table 21 like this

```{r}
mutability <- c(134, 120, 106, 102, 100,
                97,   96,  94,  93, 74,
                66,   65,  56,  56, 49,
                41,   41,  20,  20, 18)
```


Confirm that the size of this object is correct
```{r}


```



Now the names of the amino acids (which is different than the previous table we made)
```{r}

aa <- c("Asn", "Ser", "Asp", "Glu", "Ala", 
        "Thr", "Ile", "Met", "Gln", "Val",
        "His", "Arg", "Lys", "Pro", "Gly", 
        "Tyr", "Phe", "Leu", "Cys", "Trp")

```

As before, make this into a data frame; call it mut 
    `


```{r}
 mut <- data.frame(aa, mutability) 
```



Using the code from above, can you re-make table 21?  This requires square brackets, sepcifying rows, and using the cbind() function.


```{r}



```



Compare your result to Table 21.  Let me know if there are any errors.



We can merge table 21 (mut) and 22 (freqs) into a single data frame using the merge() command.  We tell the function to merge "by" the comman column, aa.

```{r}
tab.merge <- merge(freqs, mut, by = "aa")
```


We can plot the relationship between an amino acids frequency and its mutability with a scatter plot  using ggscatter
```{r}
ggscatter(data = tab.merge,
          y = "frequencies",
          x = "mutability")
```


We can add a regression line using add = "reg.line"
```{r}
ggscatter(data = tab.merge,
          y = "frequencies",
          x = "mutability",
          add = "reg.line")
```


We can calculate the correlation coefficient and a p-value suign cor.coef = TRUE.
```{r}
ggscatter(data = tab.merge,
          y = "frequencies",
          x = "mutability",
          add = "reg.line",
          cor.coef = TRUE)
```



## From observed mutations to mutation probabilities

So far we have worked up 3 things

* observed numbers of transitions between amino acids ("pams"; how frequently an aa changes to each other aa,  We revised this matrix and saved is as "A")
* observed frequencies of amino acids
* relative mutabilities (mugs; how frequently an aa changes to any other)

A useful quantity is the **probability** that a given amino acid in a sequence (say, the first aa after the start codon) is to mutate into each other amino acid, or to remain the same.  For a number of reasons, Dayhoff doesn't use a real amount of time but measure realtive to each sequence, called a "PAM."  We won't get too hung up on what this means.

The math for this mutation probability is shown on page 348 in Dayhoff et al 1978 (see 83-84 in Pevsner).  The actuall derivation is not yet totally clear to me; we'll step through the math, focusing on how to do it in R and not get bogged down in the biology.  Our main focus here will be to understand the equation and to implement the equation in R.

Dayhoff wants to calcualte a matrix (A; Figure 82) that shows the probability of an amino acid transitioning to all other aa, or staying the same.  This is the probablity of transitioning over the relative time scale of 1 PAM, so this is called a PAM1 mutation probability matrix.  Eventually we will want to use this PAM1 matrix to calculate the PAM250 matrix, which was once a popular scoring matrix for BLAST

Note that in figure 82 they multipled everything by 10000 to avoid writing decimels.  So 9867 is 98.67% or 0.9867.  This is kind of a pain; note that Pevsner doesn't do this.  Check this math below.

```{r}


```


In their notation, Dayhoff el calls the matrix of pams "A" and a single entry in the matrix Aij.  A11 is the 1st row of the 1st column.  A21 is the 2nd row, 1st column.  So, Ala's Aij value for transitioning to Arg would be the 2nd column, 1st row.

```{r}
A[2,1]
```

If this isn't clear look at the upper left part of the matrix

```{r}
tri_print(A[1:4,1:4],triangle = F)
```

We can add up all the obsereved transitions for a single amino acid, which Dayhoff calls sum(Aij), though she uses the fancy sigma symbol, like this
```{r}
sum(A[,1])
```

So, Ala was observed transitiona to another amino acid 3644 times.  For Arg we'd do this

```{r}
sum(A[,2])
```


How many times did Dayhoff observe Asp transition to another amino acid?

```{r}
# Remove the NA and add the appropriate code
# Use this result to answer the corresponding question on TopHat

pams.asp <- NA

```




Dayhoff calls the table of mutabilities (mut) "m", and each row of m is "mi".  So the mutability for Ala is m5, because Ala is in the 5th row.

```{r}
mut[5, ]
```

What code calls up Trp from the mutability matrix?

```{r}


```



Dayhoff has one other elementd in the equation, lambda (backwards h-like thing), but she doesn't give us its value.  Its a **proportionality constant** to scale things properly.  We're going to have to try to figure it out ourselves.

First let's do the easy math.  THe main part of the equation on page 348 is just:

(mj*Aij)/sum(Aij).

In words, this is multiple mutability by the number of transition Ala to Arg, then divide by the total number of times Ala transition to anything.

In reference to Ala transitioning to Arg, this is: 
"multiple the mutability of Ala (mj) by the number of times Ala transitions to Arg (Aij) and divide that by the total number of times Ala transition into anything (sum(Aij)).

These quantities are
```{r}
# m: mutability
m.Ala <- mut[5, 2]

# A.ala.arg number of transitiosn from ala to arg
A.ala.arg <- A[2,1]

# A.ala.sum: total transitions for ala
A.ala.sum <- sum(A[,1])
```

I can the multiple this out, ignoring lambda
```{r}

m.Ala*A.ala.arg/A.ala.sum

```

Let's save this to a variable x to make it easier to read

```{r}
x <- m.Ala*A.ala.arg/A.ala.sum
```


This value does not match what is shown for the Ala to Arg transtion in Dayhoff's figure 82 because

* we need to multiple by 10000
* we need to figure out lambda


The 10000 thing is annoying and we'll try out best to do the conversions in our heads.

So, doing the conversion, Dayhoff have teh Ala to Arg transition as 1, so that's really 1/10000

To get things to match we need to figure out a value for lambdas to get an answer of 1/10000.  That is, what value of lambda will result in m.Ala*A.ala.arg/A.ala.sum = 1/10000?


Ignoring the finder details of the math, the answer is:
```{r}
lambda <- (1/10000)/(m.Ala*A.ala.arg/A.ala.sum)
```

This makes lambda pretty small.
```{r}
lambda
```


So, the transition is of Ala to Arg is
```{r}
(lambda*m.Ala*A.ala.arg)/A.ala.sum
```

We can check this against Dayhoff's matrx FIgure 82 by multiplying by 10000
```{r}
10000*(lambda*m.Ala*A.ala.arg)/A.ala.sum
```


Ok, let's see if we can to this for the Ala to Asn transition, which is 4 in Dayhoff's table, or 4/10000 for us.  All the values are the same as above except A.ij, which we'll call A.ala.asn

```{r}
A.ala.asn <- A[3,1]
```

Now swap A.ala.asn in our previous math
```{r}
lambda*m.Ala*A.ala.asn/A.ala.sum
```

Now multiple by lambda
```{r}
lambda*10000*m.Ala*A.ala.asn/A.ala.sum
```

This is 3.6, and on Figure 82 for the M matrix Dayhoff has 4.  What's the difference?  Rounding.  Saving the result to an object M.ala.asn, which stands for an element of the M matrix for the ala to asn transition.

```{r}
M.ala.asn <- lambda*10000*m.Ala*A.ala.asn/A.ala.sum
```

Then we round with round()

```{r}
round(M.ala.asn,0)
```


Now, repeat these steps for the Ala to Asp transition.  This is just the next base down in the Ala column.  Make an object M.ala.Asp and round it.

```{r}
# make A.ala.Asp
# swap out the NA with the final calculation
A.ala.Asp <- NA


# Make M.ala.Asp
# swap out the NA with the final calculation
M.ala.Asp <- NA
```



So, this would be tedious to do for all 400 elements in the 20 x 20 matrix of M shown in f FIgure 82.  Lucikly R let's us do math on matrices easily.


First, I need to get my **mutability matrix** set up. Unfortunatley its not in the order I need it.  The order of the A matrix happens to be in alphabetic order by the three letter aa code, while the mutability data are in order by mutability.  The code below makes a new dataframe, mut2 that should be in the proper order; don't worry about how it works though.

```{r}
mut2 <- mut[order(mut$aa), ]
```

Now let's check to make sure that my new mut2 is the same as my matrix A.  All the response should be TRUE. (What this code is doing is comparing each pair of names and making sure they are they same)
```{r}
mut2$aa == colnames(A)
```



A key part of this equation on page 348 is the sum of all of the observed mutations, sum(Aij).  For Ala, this is the sum of the first column.  We don't want to have to do this individually for each aa, so we can use a special R function to repeat the math for us.  Don't worry how this work, but do check what the result sum.Aij looks like

```{r}
sum.Aij <- apply(A,2,sum)
```


Check out what we've done and make sure you see how it fits
```{r}
sum.Aij
```

Looking at this could help; the following code "binds" the vector sum.Aij to the matrix A (I added an extra row of dahes to make it easier to read.

```{r}
tri_print(rbind(A,rep("-",length(sum.Aij)),sum.Aij),triangle = F)
```


So, a lot is going on be we actually have all our pieces.  First, we have our original matrix A, which is the raw data.  We have our mutabilities m, which we've re-ordered to be set up in the same order as the rows of the matrix A.  We also have a new quantity, sum.Aij, which is a vector of the total number of transitions for each aa.  We also figure out what lambda is, or at least what its value is.  Now we can move from a matrix (A) that is raw data on mutations to a probability matrix.  The following code is a bit dense and I'm not going to explain it.  It uses a **for loop** to cycle through all the sets of calculations we need to do.  If you've used Python or Matlab before you might be able to figure out what's going.  Don't worry about this though, just run the code.

```{r}
# make a matrix M for mutation probabilities, the same size as 
# the original matrix A
M <- A

# make it all NAs
M[ , ] <- NA

# loop over each column of A and apply the equation 
## on page 348 by column
for(i in 1:ncol(A)){
  M[ , i] <- lambda*A[,i]*mut2$mutability[i]/sum.Aij[i]
}

```

Multiple that by 10000 to compare to Dayhoff Figure 82.
```{r}
M.10k <- M*10000
```


Let's make a temporary rouded version for viewing
```{r}
M.round <- round(M.10k, 0)
```



My values are a bit off and I'm nore sure why!  Doh!  I used an R program to scan the original data out of a PDF and I need to make sure that there were no errors (I also used the Pevsner version of the table and found 1 typo already, so perhaps there are ore)

```{r}
tri_print(M.round,triangle = F)
```



Currently we have 0 along the diagonal.  Check this

```{r}


```



According to Dayhoff the diagonal is Mij = 1-lambda*mj

m is the mutability matrix.

```{r}
Mij <- 1-lambda*mut2$mutability
```


Whatever errors I have are amplified here.  I checked the mutability values already but need to re-check them.

We can add the diagonal to our matrix M of mutation probablities like this:
```{r}
diag(M) <- Mij
```


So, what we have here is a PAM1 mutation probability matrix.  See Pevsner and Dayhoff for the precise intepretation of this.  



## From PAM1 to PAM250

As discussed by Dayhoff and Pevsner PAM1 is usually converted to PAM250.  PAM1 represents transition probabilities over a short period of time (1 PAM) while more practically a longer period of time is more useful, like 250 PAMS.  (Don't worry about what a PAM is; is a relativel measure of evolutionary time.)


Converting from PAM1 to PAM250 is each.  We just multiple the matrix by itself.  Basically we multiple each element of the matrix by iteslf.

PAM2 would look like this, where %*% is a special tool for multiplying matices by each other.

```{r}
PAM2 <- M%*%M
```

Below is the upper left hand corner of PAM2

```{r}
tri_print(round(PAM2[1:4,1:4],3),triangle = F)
```

Challenge: can you figure out how to confirm the value for Ala to Ala?

PAM10 would be like this

```{r}
PAM10 <- M%*%M%*%M%*%M%*%M%*%M%*%M%*%M%*%M%*%M
```

What do you notice about the numbers?  What happens when you multiple fractional values by each other?
```{r}
tri_print(round(PAM10[1:4,1:4],3),triangle = F)
```


For the Ala->Ala transition in the upper left corner what this represents is Ala remaining Ala over the course of 10 time steps (10 PAMs).  As you increase the number of time steps, it becomes less and less likely that Ala won't change, so this number goes down.

Getting to PAM250 would be a pain.  Luckily we use a package to do the matrix equivalent of raising the matrix to the 10th power.


```{r}
install.packages("matrixcalc")
library(matrixcalc)
```

NOTE: Pevsner has a box showing how to do this and is totally wrong!

```{r}
PAM10.2 <- matrix.power(M, 10)
```



We can confirm this like this.   All the values are TRUE because M*M*M*M*M*M*M*M*M*M and matrix.power(M, 10) produce the same result in every single cell.
```{r}
PAM10 == PAM10.2


```


Let's make PAM250.  We'll call it M250 because we're still working with a mutation probablity matrix and not a scoring matrix.

```{r}

M250 <- matrix.power(M, 250)

```


Let's look at the upper left part of the matrix.  I've multipled it by 100 so it looks more like Dayhoff.  Still off by a bit
```{r}
tri_print(round(M250[1:4,1:4],2)*100,triangle = F)
```


There's  specific interpretation of this matrix.  See Pevsner and Dayhoff for how to interpret these values.

## Making the final matrix

So far we're still working in terms of probabilities, eg, what the probablity Ala stays Ala over a "distance" of 250 PAMS?  What's the probablity that Ala transitions to Arg over 250 PAMs?

We need a bit more math to get to the scoring matrix.  We'll gloss over the why of the math in order to just pratice R.

On page 351 Dayhoff has an equation 
Rij = Mij/fi

This is a "relatedness odds matrix", which is a fancier name for an alignment scoring matrix.  If you know gambling or statistics you might know about odds.  Don't worry about it for now.

M is probabilty matrix, eg M250.  f is the relative frequency of each amino acids (freqs).  Let's do the math.  If you want a challenge, try to figure out how to do it yourself.  Key steop: putting the frequencies in the right order!

```{r}
freqs2 <- freqs[order(freqs$aa), ]
```


We nee a for loop again.  Just run this unless you want to root aroudn in teh details
```{r}
# make a matrix M for mutation probabilities, the same size as 
# the original matrix A
R250 <- M250

# make it all NAs
R250[ , ] <- NA

# loop over each column of A and apply the equation 
## on page 348 by column
for(i in 1:ncol(M250)){
  R250[ , i] <- M250[,i]/(freqs2$frequencies[i])
}
```

R is a matrix of odds ratios.  Don't worry about how to interpret them; its still not a scoring matrix.  To get to the scoring matrix, we need to do one more mathy thing invovling log10, and multiplying by 10.  Dayhoff doesn't cover all of this but see equation 3.4 in Pevsner (pag 89)
```{r}
PAM250 <- 10*log10(R250)
```


Let's round that off

```{r}
PAM250.round <- round(R250, 0)
```

